# Model Selection and Interpretation

## Covariate Adjustment

Back to example 4.3 to illustrate the importance of covariate adjustment

```
### won't run. the data is simulated. see chapter 4
### beta ia positive, increasing the risk
coxph(Surv(ttAll, status)~trt)

### beta changed to negative, decreaing the risk
coxph(Surv(ttAll, status)~trt+strata(genotype))
coxph(Surv(ttAll, status)~trt+genotype)
```

## Categorical and Continuous Covariates

indicator or dummy variable: values 0 or 1

categorical variables with three or more values, we need multiple dummy variables. one level is selected as a reference level, arbitrarily or driven by the goals of the research project.

Let us say we have a set of k covariates, we can write the model

$$log(\psi_i)=z_{1i}\beta_1+z_{2i}\beta_2+...+z_{ki}\beta_k$$

- the parameter $\beta_j$ is the log hazard ratio
- for continuous covariates, it represents the effect of a unit change in the covariate
- for dummy variables, it represents the effect of the corresponding level as compared to the reference covariate

- non-linear: consider fit a transformation of a covariate, e.g. logarithmic or square root; or "discretize", such as "Age<50"
- interaction

Comparing with linear/logistic regression,

- time-related variables: here we only considering fixed at the begining of the trial (baseline)
- no intercept: "absorbed into the baseline hazard"

Example 6.1

```{r}
race <- factor(c("black", "black", "white", "white", "other",
"other"))
age <- c(48, 52, 87, 82, 67, 53)

model.matrix(~ race + age)[,-1]

### taking white as the reference level
race <- relevel(race, ref="white")
model.matrix(~ race + age)[,-1]

## include an interaction: 
## the product between race (first two columns) and age (third column)
model.matrix(~ race + age + race:age)[,-1]
```

Example 6.2

```{r}
### simulate the data
## age from 40 to 80
age <- runif(n=60, min=40, max=80)
## race with three levels
race <- factor(c(rep("white", 20), rep("black", 20), rep("other", 20)))
race <- relevel(race, ref="white")

### simulate the survival by a exponential distribution
### with the rate parameter
log.rate.vec = -4.5+c(rep(0,20),rep(1,20),rep(2,20))+age*0.05
### and no censoring for the survival status
tt = rexp(n=60, rate=exp(log.rate.vec))
status = rep(1,60)

### now fit a cox proportional hazards model
library(survival)
result.cox = coxph(Surv(tt, status)~race+age)
summary(result.cox)
```

## Hypothesis Testing for Nested Models

Nested models: the covariates of one model is a subset of the covariates in the other

We can compare nested models using a (partial) likelihood ratio test

```{r}
library(asaur)
attach(pharmacoSmoking)
## model A: ageGroup4
modelA.coxph = coxph(Surv(ttr, relapse)~ageGroup4)
modelA.coxph

## model B: employment
modelB.coxph = coxph(Surv(ttr, relapse)~employment)
modelB.coxph

## model C: ageGroup4+employment
modelC.coxph = coxph(Surv(ttr, relapse)~ageGroup4+employment)
modelC.coxph

### log-likelihood
logLik(modelA.coxph)
logLik(modelB.coxph)
logLik(modelC.coxph)

### model A vs model C
pchisq(2*(as.numeric(logLik(modelC.coxph))-as.numeric(logLik(modelA.coxph))), df=5-3, lower.tail=F)

### model B vs model C
pchisq(2*(as.numeric(logLik(modelC.coxph))-as.numeric(logLik(modelB.coxph))), df=5-2, lower.tail=F)
```

## The Akaike Information Criterion for Comparing Non-nested Models

## Including Smooth Estimates of Continuous Covariates in a Survival Model

